---
title: "compare_eigenvectors"
output: html_document
---

Alexey Larionov, 07Feb2017  

# Summary

Compare eigenvectors calculated by R-script and by EIGENSTRAT-tool  
- using wecare only dataset (480 cases x 226,722 variants)  
- w/o removing outliers (EIGENSTRAT outliermode 2)  
- using Price 2006 normalising (EIGENSTRAT altnormstyle NO)  

# start_section
echo=FALSE, results='hide'
```{r start_section, echo=FALSE, results='hide'}

# Time stamp
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"

# Required libraries
library(dplyr)
library(ggplot2)
library(plotly)

```

# read_data
echo=FALSE, results='hide'
```{r read_data, echo=FALSE, results='hide'}

# --- r-calculated data --- #

load(paste(interim_data_folder, "r05_calculate_egenvectors_wecare_only_jan2017.RData", sep="/"))

rm(wecare_exac.df, wecare_genotypes.mx, wecare_kgen.df, wecare_nfe_exac.df, wecare_nfe_genotypes.mx, wecare_nfe_kgen.df, wecare_nfe_phenotypes.df, wecare_nfe_variants.df, wecare_phenotypes.df, wecare_variants.df)

# eigenvectors
r_evectors.df <- as.data.frame(wecare_eigen$vectors[,1:10])

colnames(r_evectors.df) <- c("r_ev1", "r_ev2", "r_ev3", "r_ev4", "r_ev5", "r_ev6", "r_ev7", "r_ev8", "r_ev9", "r_ev10")
str(r_evectors.df)

# Common sence checks (using properties of eigenvectors)
apply(r_evectors.df,2,sum) # should be close to zeros
apply(r_evectors.df,2,function(x){sum(x^2)}) # should be close to ones

# eigenvalues
r_evalues.df <- as.data.frame(wecare_eigen$values)
colnames(r_evalues.df) <- "evals"

# --- es-calculated data --- #

es_evectors_file="/scratch/medgen/scripts/wecare_stat_01.17/interim_data/u01_wecare_only_480_226k_eigenstrat_no_outliers/u01_wecare_only_480_226k_eigenstrat_no_outliers.evec"

es_evectors.df <- read.table(es_evectors_file)

colnames(es_evectors.df) <- c("wes_id", "es_ev1", "es_ev2", "es_ev3", "es_ev4", "es_ev5", "es_ev6", "es_ev7", "es_ev8", "es_ev9", "es_ev10", "group")

str(es_evectors.df)

# Common sence checks (using properties of eigenvectors)
apply(es_evectors.df[,2:11],2,sum) # should be close to zeros
apply(es_evectors.df[,2:11],2,function(x){sum(x^2)}) # should be close to ones

es_evalues_file <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data/u01_wecare_only_480_226k_eigenstrat_no_outliers/u01_wecare_only_480_226k_eigenstrat_no_outliers.eval"

es_evalues.df <- read.table(es_evalues_file)
colnames(es_evalues.df) <- "evals"
str(es_evalues.df)

rm(es_evectors_file, es_evalues_file)

```

# compare_evalues

```{r compare_evalues}

plot(r_evalues.df$evals, 
     main="R-script eigenvalues", xlab="Eigenvectors", ylab="Eigenvalues")

plot(es_evalues.df$evals, 
     main="EIGENSTRAT-tool eigenvalues", xlab="Eigenvectors", ylab="Eigenvalues")

plot(es_evalues.df$evals, r_evalues.df$evals, 
     main="Eigenvalues: R-script vs EIGENSTRAT-tool", xlab="EIGENSTRAT-tool", ylab="R-script")
     abline(c(0,0),c(1,1), col="red")

```

# compare_evectors

```{r compare_evectors}

# Prepare dataframe
joined_evectors.df <- cbind(es_evectors.df, r_evectors.df)

# Prepare colour scale
colours <- c("UBC" = "BLUE", "CBC" = "RED")
userColourScale <- scale_colour_manual(values=colours)

# --- 1st eigenvectors --- #

# Prepare plot object
g <- ggplot(joined_evectors.df, aes(es_ev1, r_ev1)) +
  geom_point(aes(colour=group, fill=group, text=wes_id)) + 
  labs(title="1st eigenvectors: R vs EIGENSTRAT") +
  geom_abline(colour = "brown", linetype = 2) + 
  userColourScale

# Render interactive plot
ggplotly(g)

# --- 2nd eigenvectors --- #

# Prepare plot object
g <- ggplot(joined_evectors.df, aes(es_ev2, r_ev2)) +
  geom_point(aes(colour=group, fill=group, text=wes_id)) + 
  labs(title="2nd eigenvectors: R vs EIGENSTRAT") +
  geom_abline(colour = "brown", linetype = 2) + 
  userColourScale

# Render interactive plot
ggplotly(g)

# --- 3rd eigenvectors --- #

# Prepare plot object
g <- ggplot(joined_evectors.df, aes(es_ev3, r_ev3)) +
  geom_point(aes(colour=group, fill=group, text=wes_id)) + 
  labs(title="3rd eigenvectors: R vs EIGENSTRAT") +
  geom_abline(slope = -1, colour = "brown", linetype = 2) + 
  userColourScale

# Render interactive plot
ggplotly(g)

```

# final_section

```{r final_section, results='hold'}

ls()
sessionInfo()
Sys.time()

```
