---
title: "consolidate_phenotype_data_wecare_jan2017"
output: html_document
---

started: Alexey Larionov, 01Mar2016  
last updated: Alexey Larionov, 26Jan2017

# Summary

This step prepares phenotype data for downstream analysis. 
It selects and merges data from various, partially duplicated and incomplete source files: 
(covar, demographics, samples, phenotypes_update, BRCA1/2-PALB2 carriers)

The data selection is based on the data assessment done during previous versions of analysis (feb, aug, sep, nov 2016).

###Notes:  

1) Remember to impute one truly missed hormones datapoint (P3_H06, later)

2) Despite the fact that **phenotypes.df** and **gt.mx** include the same number of samples, 
 *wes_IDs* in **phenotypes.df** are not yet syncronised with colnames of **gt.mx**:

- gt.mx has some saqmples names mofifyed or missed (e.g. missing P1_B08, containing P1_E07 and P1_E07_Copy etc)

- order of samples in gt.mx colnames was set by GATK during combining g.vcfs. 
This order is different from what was was used in samples.txt file, which propogated to phenotypes.df  

3) nfe cases are coded as -1 in cc field

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"

# Do NOT clean-up at this stage
#rm(list=ls()) # this would remove params object!
#graphics.off()

# Required libraries
library(dplyr) # for piping, filter, select etc
library(reshape2) # for melt()
library(ggplot2)

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r02_filter_genotypes_and_variants_jan2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(gt.mx)
class(gt.mx)
gt.mx[1:5,1:5]

dim(covar.df)
str(covar.df)
covar.df[1:5,1:5]

dim(samples.df)
str(samples.df)
samples.df[1:5,]

dim(demographics.df)
str(demographics.df)
demographics.df[1:5,1:5]

dim(phenotypes_update.df)
str(phenotypes_update.df)
phenotypes_update.df[1:5,1:5]

dim(BRCA1_BRCA2_PALB2_cases.df)
str(BRCA1_BRCA2_PALB2_cases.df)
BRCA1_BRCA2_PALB2_cases.df[1:5,1:5]

dim(vv.df)
str(vv.df)
vv.df[1:5,1:5]

dim(kgen.df)
str(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
str(exac.df)
exac.df[1:5,1:5]

# Check consistence of rownames
sum(rownames(gt.mx) != rownames(vv.df))
sum(rownames(gt.mx) != rownames(kgen.df))
sum(rownames(gt.mx) != rownames(exac.df))

```

# select_variants_annotations

Select variants annotations (based on the previous data assessments)
This has nothing to do with phenotypes, but still needs to be done somewhere.

```{r select_variants_annotations}

variants.df <- 
  vv.df %>% 
  select(SplitVarID, SYMBOL, TYPE, CHROM, POS, REF, ALT, AC, AF, AN, 
         Consequence, SIFT_call, SIFT_score, PolyPhen_call, PolyPhen_score, CLIN_SIG, 
         cDNA_position, CDS_position, Codons, Protein_position, Amino_acids, 
         Existing_variation, Multiallelic)

dim(variants.df)
str(variants.df)
variants.df[1:5,1:5]

rm(vv.df)

```

# reshape_covar_table

Select data from covar table (based on the previous data assessments)

```{r reshape_covar_table}

# Keep only selected annotations
covar.df <- covar.df %>% 
  select(labid, setno, cc, sub_dx_age, chemo, hormone, XRTBreast, dose, 
         eig1_covar = Eigen_1, 
         eig2_covar = Eigen_2, 
         eig3_covar = Eigen_3)

# change XRTBreast to xrt
"xrt" -> colnames(covar.df)[7]

# Check result
dim(covar.df)
str(covar.df)
covar.df[1:5,1:5]

```

# reshape_demographics_table

Select data from demographics table (based on the previous data assessments)

## odd_cases

Removing odd cases (based on the previous data assessments)

```{r odd_cases}

# Outlook
dim(demographics.df)
colnames(demographics.df)
demographics.df[1:10,1:5]

# List of odd cases
odd_cases=c("201558","201921","202698","21IAno","387460",
            "389192","58IR1+","60IA1+","92IRno","98IAno")

# Show odd cases
odd_cases.df <- demographics.df %>% filter(Subject_ID %in% odd_cases)
odd_cases.df[1:5]

# Cases to preserve:
# 201558, 201921, 387460, 389192 - added in dec 2016
# 202698 - present in covar

# Add missed labid to the cases, which will be preserved
as.vector(demographics.df$labid.x) -> demographics.df$labid.x
"id201558" -> demographics.df[demographics.df$Subject_ID == "201558", "labid.x"]
"id201921" -> demographics.df[demographics.df$Subject_ID == "201921", "labid.x"]
"id387460" -> demographics.df[demographics.df$Subject_ID == "387460", "labid.x"]
"id389192" -> demographics.df[demographics.df$Subject_ID == "389192", "labid.x"]
"id202698" -> demographics.df[demographics.df$Subject_ID == "202698", "labid.x"]

# Cases to exclude: odd cases IDs and data
cases_to_exclude=c("21IAno","58IR1+","60IA1+","92IRno","98IAno")

demographics.df <- demographics.df %>% filter(! Subject_ID %in% cases_to_exclude)
dim(demographics.df)
demographics.df[1:10,1:5]

rm(odd_cases, odd_cases.df, cases_to_exclude)

```

## duplicated_cases - not needed to remove? - it will be filtered by "filter" field

Removing duplicates (based on the previous data assessments)  

259643: P4_F01, P5_D08  
272715: P4_F02, P5_D07  

The demographic data in duplicates were identical. 
The duplicates removal is necessary for proper merging with covar. 

```{r duplicated_cases}

# Remove the duplicated cases (except the 1st duplicate)
duplicated_cases_bool <- duplicated(demographics.df$Subject_ID) # boolean, excluding position of the first duplicate
demographics.df <- demographics.df[!duplicated_cases_bool,]

# Explore result
dim(demographics.df)
demographics.df[1:10,1:5]

# Clean-up
rm(duplicated_cases_bool)

```

## select_demographics_data

Selecting data (based on the previous data assessments)

```{r select_demographics_data}

# Select columns
demographics.df <- demographics.df %>% select(
  labid = labid.x,
  cc = cc.x,
  setno = setno.x,
  family_history = family_history.y,
  sub_dx_age = sub_dx_age.x,
  refage = refage,
  rstime = rstime,
  eig1_demo = Eigen_1.x,
  eig2_demo = Eigen_2.x,
  eig3_demo = Eigen_3.x,
  stage_fd = stage_fd,
  er1 = er1,
  pr1_cat = pr1_cat,
  histo1_cat = histo1_cat,
  hormone = hormone,
  CMF = CMF,
  XRTBrCHAR = XRTBrCHAR,
  dose_caseloc = dose_caseloc,
  rh_age_menarche = rh_age_menarche,
  age_1fftp_fd = age_1fftp_fd, 
  age_menopause_1yrbf_fd = age_menopause_1yrbf_fd,
  num_preg = num_preg,
  BMI_age18 = BMI_age18,
  BMI_dx = BMI_dx,
  BMI_ref = BMI_ref)

# Recode categorical pr to NA, 0 and 1
demographics.df <- demographics.df %>% mutate(
  pr1 = ifelse(pr1_cat=="negative", 0,
        ifelse(pr1_cat=="positive", 1,
        NA)))

# Substitute zero age of menarche to NA
demographics.df <- demographics.df %>% mutate(
  age_menarche = ifelse(rh_age_menarche==0, NA, rh_age_menarche))

# Remove initial fields
demographics.df <- demographics.df %>% select(-rh_age_menarche, -pr1_cat)

```

# merge_demographics_and_covar

```{r merge_demographics_and_covar}

# Check data frames
dim(covar.df)
dim(demographics.df)
colnames(covar.df)
colnames(demographics.df)

# Prepare key for joining
covar.df$labid <- as.vector(covar.df$labid)
demographics.df$labid <- as.vector(demographics.df$labid)

# Join data frames
pheno.df <- left_join(covar.df, demographics.df, by="labid")
dim(pheno.df)
colnames(pheno.df)

# Eigenvectors are the same (a common-sense check)
sum(pheno.df$eig1_demo != pheno.df$eig1_covar, na.rm=TRUE)
sum(pheno.df$eig2_demo != pheno.df$eig2_covar, na.rm=TRUE)
sum(pheno.df$eig3_demo != pheno.df$eig3_covar, na.rm=TRUE)

# Select columns
pheno.df <- pheno.df %>% 
  select(
    gwas_id = labid, 
    cc = cc.x, 
    setno = setno.x,
    family_history = family_history,
    age_dx = sub_dx_age.x, 
    age_ref = refage, 
    rstime = rstime,
    eig1_gwas = eig1_covar, 
    eig2_gwas = eig2_covar, 
    eig3_gwas = eig3_covar, 
    stage = stage_fd, 
    er = er1, 
    pr = pr1, 
    hist_cat = histo1_cat, 
    hormone = hormone.x, 
    chemo_cat = CMF, 
    br_xray = XRTBrCHAR, 
    br_xray_dose = dose_caseloc, 
    age_menarche = age_menarche, 
    age_1st_ftp = age_1fftp_fd, 
    age_menopause = age_menopause_1yrbf_fd, 
    num_preg = num_preg, 
    bmi_age18 = BMI_age18, 
    bmi_dx = BMI_dx, 
    bmi_ref = BMI_ref)

dim(pheno.df)
colnames(pheno.df)

rm(covar.df, demographics.df)

```

# merge_pheno_and_samples

This step adds wes_IDs and data to filter samples basing on wes performance. 

Importantly: wes_IDs are not yet syncronised with colnames of gt.mx:

- gt.mx has some saqmples names mofifyed or missed (e.g. missing P1_B08, containing P1_E07 and P1_E07_Copy etc)

- order of samples in gt.mx colnames was set by GATK during combining g.vcfs. 
This order is different from what was was used in samples.txt file, which propogated to phenotypes.df

```{r merge_pheno_and_samples}

# Prepare key for joining
samples.df$gwas_id <- as.vector(samples.df$gwas_id)

# Join data frames
phenotypes.df <- full_join(samples.df, pheno.df, by="gwas_id")

# Check results
dim(phenotypes.df)
colnames(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:10, 1:7]

# reformat_phenotypes gwas and wes ids ...
str(phenotypes.df)

as.vector(phenotypes.df$wes_id) -> phenotypes.df$wes_id
as.vector(phenotypes.df$merged_id) -> phenotypes.df$merged_id
as.vector(phenotypes.df$filter) -> phenotypes.df$filter
as.vector(phenotypes.df$hist_cat) -> phenotypes.df$hist_cat
NA -> phenotypes.df$hist_cat[phenotypes.df$hist_cat == "unknown"]
as.vector(phenotypes.df$chemo_cat) -> phenotypes.df$chemo_cat

str(phenotypes.df)

# Clean-up
rm(samples.df, pheno.df)

```

# add_new_pheno_data

Remember to impute one truly missed hormones datapoint (P3_H06, later)

```{r add_new_pheno_data}

# For each sample with phenotypes update
for(i in 1:nrow(phenotypes_update.df)){
  
  # Read updated data
  gwas_id_upd <- phenotypes_update.df[i,"gwas_id"]
  wes_id_upd <- phenotypes_update.df[i,"wes_id"]
  cc_upd <- phenotypes_update.df[i,"cc"]
  setno_upd <- phenotypes_update.df[i,"setno"]
  family_history_upd <- phenotypes_update.df[i,"family_history"]
  age_dx_upd <- phenotypes_update.df[i,"age_dx"]
  age_ref_upd <- phenotypes_update.df[i,"age_ref"]
  rstime_upd <- phenotypes_update.df[i,"rstime"]
  eig1_upd <- phenotypes_update.df[i,"Eigen_1"]
  eig2_upd <- phenotypes_update.df[i,"Eigen_2"]
  eig3_upd <- phenotypes_update.df[i,"Eigen_3"]
  stage_upd <- phenotypes_update.df[i,"stage"]
  er_upd <- phenotypes_update.df[i,"er1"]
  pr_upd <- phenotypes_update.df[i,"pr1"]
  hist_cat_upd <- phenotypes_update.df[i,"hist_cat"]
  hormone_upd <- phenotypes_update.df[i,"hormone"]
  chemo_cat_upd <- phenotypes_update.df[i,"chemo_cat"]
  br_xray_upd <- phenotypes_update.df[i,"br_xray"]
  br_xray_dose_upd <- phenotypes_update.df[i,"br_xray_dose"]
  num_preg_upd <- phenotypes_update.df[i,"num_preg"]

  # Progress report
  print(paste("---", wes_id_upd, "---"))

  # Get rownum in phenotypes data frame
  j <- phenotypes.df[,"wes_id"] == wes_id_upd
  
  # Stop with error if count of wes_id in initial data != 1
  if(sum(j) != 1){
    print("Error with wes_id")
    print("Script terminated") 
    stop()
  }

  # --- Check and update gwas_id --- #
  
  # if updated data is NA
  if(is.na(gwas_id_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"gwas_id"])){
      print("Error with gwas_id")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"gwas_id"])){
      gwas_id_upd -> phenotypes.df[j,"gwas_id"]
      print(paste("Updated gwas_id"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"gwas_id"] != gwas_id_upd){
      print("Error with gwas_id")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update cc --- #
  
  # if updated data is NA
  if(is.na(cc_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"cc"])){
      print("Error with cc")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"cc"])){
      cc_upd -> phenotypes.df[j,"cc"]
      print("Updated cc")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"cc"] != cc_upd){
      print("Error with cc")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update setno --- #
  
  # if updated data is NA
  if(is.na(setno_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"setno"])){
      print("Error with setno")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"setno"])){
      setno_upd -> phenotypes.df[j,"setno"]
      print("Updated setno")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"setno"] != setno_upd){
      print("Error with setno")
      print("Script terminated") 
      stop()
    }
  }

  # --- Check and update family_history --- #
  
  # if updated data is NA
  if(is.na(family_history_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"family_history"])){
      print("Error with family_history")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"family_history"])){
      family_history_upd -> phenotypes.df[j,"family_history"]
      print("Updated family_history")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"family_history"] != family_history_upd){
      print("Error with family_history")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update age_dx --- #
  
  # if updated data is NA
  if(is.na(age_dx_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"age_dx"])){
      print("Error with age_dx")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"age_dx"])){
      age_dx_upd -> phenotypes.df[j,"age_dx"]
      print("Updated age_dx")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"age_dx"] != age_dx_upd){
      print("Error with age_dx")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update age_ref --- #
  
  # if updated data is NA
  if(is.na(age_ref_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"age_ref"])){
      print("Error with age_ref")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"age_ref"])){
      age_ref_upd -> phenotypes.df[j,"age_ref"]
      print("Updated age_ref")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"age_ref"] != age_ref_upd){
      print("Error with age_ref")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update rstime --- #
  
  # if updated data is NA
  if(is.na(rstime_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"rstime"])){
      print("Error with rstime")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"rstime"])){
      rstime_upd -> phenotypes.df[j,"rstime"]
      print("Updated rstime")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"rstime"] != rstime_upd){
      print("Error with rstime")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update eig1 --- #
  
  # if updated data is NA
  if(is.na(eig1_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"eig1_gwas"])){
      print("Error with eig1")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"eig1_gwas"])){
      eig1_upd -> phenotypes.df[j,"eig1_gwas"]
      print("Updated eig1")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"eig1_gwas"] != eig1_upd){
      print("Error with eig1")
      print("Script terminated") 
      stop()
    }
  }

  # --- Check and update eig2 --- #
  
  # if updated data is NA
  if(is.na(eig2_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"eig2_gwas"])){
      print("Error with eig2")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"eig2_gwas"])){
      eig2_upd -> phenotypes.df[j,"eig2_gwas"]
      print("Updated eig2")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"eig2_gwas"] != eig2_upd){
      print("Error with eig2")
      print("Script terminated") 
      stop()
    }
  }

  # --- Check and update eig3 --- #
  
  # if updated data is NA
  if(is.na(eig3_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"eig3_gwas"])){
      print("Error with eig3")
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"eig3_gwas"])){
      eig3_upd -> phenotypes.df[j,"eig3_gwas"]
      print("Updated eig3")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"eig3_gwas"] != eig3_upd){
      print("Error with eig3")
      print("Script terminated") 
      stop()
    }
  }

  # --- Check and update stage --- #
  
  # if updated data is NA
  if(is.na(stage_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"stage"])){
      print(paste("Error with stage"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"stage"])){
      stage_upd -> phenotypes.df[j,"stage"]
      print("Updated stage")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"stage"] != stage_upd){
      print("Error with stage")
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update er --- #
  
  # if updated data is NA
  if(is.na(er_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"er"])){
      print(paste("Error with er"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"er"])){
      er_upd -> phenotypes.df[j,"er"]
      print("Updated er")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"er"] != er_upd){
      print(paste("Error with er"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update pr --- #
  
  # if updated data is NA
  if(is.na(pr_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"pr"])){
      print(paste("Error with pr"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"pr"])){
      pr_upd -> phenotypes.df[j,"pr"]
      print("Updated pr")
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"pr"] != pr_upd){
      print(paste("Error with pr"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update hist_cat --- #
  
  # if updated data is NA
  if(is.na(hist_cat_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"hist_cat"])){
      print(paste("Error with hist_cat"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"hist_cat"])){
      hist_cat_upd -> phenotypes.df[j,"hist_cat"]
      print(paste("Updated hist_cat"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"hist_cat"] != hist_cat_upd){
      print(paste("Error with hist_cat"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update hormone --- #
  
  # if updated data is NA
  if(is.na(hormone_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"hormone"])){
      print(paste("Error with hormone"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"hormone"])){
      hormone_upd -> phenotypes.df[j,"hormone"]
      print(paste("Updated hormone"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"hormone"] != hormone_upd){
      print(paste("Error with hormone"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update chemo_cat --- #
  
  # if updated data is NA
  if(is.na(chemo_cat_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"chemo_cat"])){
      print(paste("Error with chemo_cat"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"chemo_cat"])){
      chemo_cat_upd -> phenotypes.df[j,"chemo_cat"]
      print(paste("Updated chemo_cat"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"chemo_cat"] != chemo_cat_upd){
      print(paste("Error with chemo_cat"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update br_xray --- #
  
  # if updated data is NA
  if(is.na(br_xray_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"br_xray"])){
      print(paste("Error with br_xray"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"br_xray"])){
      br_xray_upd -> phenotypes.df[j,"br_xray"]
      print(paste("Updated br_xray"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"br_xray"] != br_xray_upd){
      print(paste("Error with br_xray"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update br_xray_dose --- #
  
  # if updated data is NA
  if(is.na(br_xray_dose_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"br_xray_dose"])){
      print(paste("Error with br_xray_dose"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"br_xray_dose"])){
      br_xray_dose_upd -> phenotypes.df[j,"br_xray_dose"]
      print(paste("Updated br_xray_dose"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"br_xray_dose"] != br_xray_dose_upd){
      print(paste("Error with br_xray_dose"))
      print("Script terminated") 
      stop()
    }
  }
  
  # --- Check and update num_preg --- #
  
  # if updated data is NA
  if(is.na(num_preg_upd)){ 
    
    # Stop with error if initial data is not NA
    if(!is.na(phenotypes.df[j,"num_preg"])){
      print(paste("Error with num_preg"))
      print("Script terminated") 
      stop()
    }
  
  # If updated data is not NA
  }else{

    # Update if initial data is NA
    if(is.na(phenotypes.df[j,"num_preg"])){
      num_preg_upd -> phenotypes.df[j,"num_preg"]
      print(paste("Updated num_preg"))
    }
    
    # Stop with error if the initial data is different from the updated
    if(phenotypes.df[j,"num_preg"] != num_preg_upd){
      print(paste("Error with num_preg"))
      print("Script terminated") 
      stop()
    }
  }
} # next sample

# correct filter field
updated_samples <- c("P1_C02", "P1_C04", "P1_C06", "P1_C11", "P1_E10", 
                     "P1_F01", "P1_F10", "P1_H05", "P3_H06", "P3_H07", 
                     "P3_H08", "P3_H09", "P3_H10")

sum(phenotypes.df$filter == "not_annotated")
"pass" -> phenotypes.df[ phenotypes.df$wes_id %in% updated_samples ,"filter"]
sum(phenotypes.df$filter == "not_annotated")

# Clean-up
rm(i, j, gwas_id_upd, wes_id_upd, cc_upd, setno_upd, family_history_upd,
  age_dx_upd, age_ref_upd, rstime_upd, eig1_upd, eig2_upd, eig3_upd, 
  stage_upd, er_upd, pr_upd, hist_cat_upd, hormone_upd, chemo_cat_upd, 
  br_xray_upd, br_xray_dose_upd, num_preg_upd, phenotypes_update.df, 
  updated_samples)

```

# add_information_about_BRCA1_BRCA2_PALB2_carriers

```{r add_information_about_BRCA1_BRCA2_PALB2_carriers}

# Mutation carriers
BRCA1_carriers <- BRCA1_BRCA2_PALB2_cases.df %>% filter(SYMBOL=="BRCA1") %>% select(Cases_wes_id)
BRCA1_carriers <- as.vector(BRCA1_carriers$Cases_wes_id)
BRCA1_carriers

BRCA2_carriers <- BRCA1_BRCA2_PALB2_cases.df %>% filter(SYMBOL=="BRCA2") %>% select(Cases_wes_id)
BRCA2_carriers <- as.vector(BRCA2_carriers$Cases_wes_id)
BRCA2_carriers

PALB2_carriers <- BRCA1_BRCA2_PALB2_cases.df %>% filter(SYMBOL=="PALB2") %>% select(Cases_wes_id)
PALB2_carriers <- as.vector(PALB2_carriers$Cases_wes_id)
PALB2_carriers

# Update phenotypes filter column
rownames(phenotypes.df) <- phenotypes.df$wes_id

phenotypes.df[BRCA1_carriers,"filter"]
"BRCA1_carrier" -> phenotypes.df[BRCA1_carriers,"filter"]
phenotypes.df[BRCA1_carriers,"filter"]

phenotypes.df[BRCA2_carriers,"filter"]
"BRCA2_carrier" -> phenotypes.df[BRCA2_carriers,"filter"]
phenotypes.df[BRCA2_carriers,"filter"]

phenotypes.df[PALB2_carriers,"filter"]
"PALB2_carrier" -> phenotypes.df[PALB2_carriers,"filter"]
phenotypes.df[PALB2_carriers,"filter"]

rm(BRCA1_carriers, BRCA2_carriers, PALB2_carriers, BRCA1_BRCA2_PALB2_cases.df)

```

# write_wecare_phenotypes_table

```{r write_wecare_phenotypes_table}

write.table(phenotypes.df, file=paste(interim_data_folder, "wecare_phenotypes.txt", sep="/"), quote=FALSE, sep="\t")

```

# explore_cases

of 512 sequenced cases:  
- 19 failed sequencing (mixed during the library prep, low concordance with gwas, some of those were also non-annotated and BRCA1/2 carriers)  
- 2 were intentional duplicates  
- 4 BRCA1 carriers (3 intentional)  
- 4 BRCA2 carriers  
- 3 PALB2 carriers

```{r explore_cases}

summary(as.factor(phenotypes.df$filter))

```

# explore_consistency_of_some_clinical_data

Apparently, 11 ER-negatives were treated with hormones; 2 of ththem ER/PR-negatives ...

```{r explore_consistency_of_some_clinical_data}

phenotypes.df %>% 
  filter(er==0 & hormone == 1) %>% 
  select(wes_id, gwas_id, er, pr, hormone)

```

# add_nfe_to_phenotypes_table

Encoding:  
- ID to wes_id and rownames  
- "-1"" to cc  
- "pass"" to filter  

```{r add_nfe_to_phenotypes_table}

nfe_phenotypes.df <- as.data.frame(matrix(rep(NA,198*28), nrow=198))
colnames(nfe_phenotypes.df) <- colnames(phenotypes.df)

nfe_ids <- colnames(gt.mx)[1:198]
nfe_phenotypes.df$wes_id <- nfe_ids
rownames(nfe_phenotypes.df) <- nfe_ids
"pass" -> nfe_phenotypes.df$filter
-1 -> nfe_phenotypes.df$cc

phenotypes.df <- rbind(nfe_phenotypes.df, phenotypes.df)

summary(as.factor(phenotypes.df$filter))

dim(gt.mx)
dim(phenotypes.df)
dim(variants.df)

rm(nfe_ids, nfe_phenotypes.df)

```

# data_summary

```{r data_summary}

ls()

dim(gt.mx)
class(gt.mx)
gt.mx[1:5,1:5]

dim(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:5,1:5]

dim(variants.df)
colnames(variants.df)
variants.df[1:5,1:5]

dim(kgen.df)
colnames(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
colnames(exac.df)
exac.df[1:5,1:5]

# Check consistency of rownames
sum(rownames(gt.mx) != rownames(variants.df))
sum(rownames(gt.mx) != rownames(kgen.df))
sum(rownames(gt.mx) != rownames(exac.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r03_consolidate_phenotype_data_jan2016.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
