---
title: "calculate_eigenvectors_wecare_only_jan2017_explore"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 01Mar2017

# Summary

This script calculates eigenvectors from common variants in wecare only dataset and compares it with eigenvectors calculates on all variants.

Exploring this direction I desided to use a conventional approach of calculating eigenvectors from all vsriants, not from the common variants only.  

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"

# Required libraries
library(dplyr)
library(ggplot2)
library(plotly)

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r05a_calculate_egenvectors_wecare_only_jan2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_genotypes.mx)
class(wecare_genotypes.mx)
wecare_genotypes.mx[1:5,1:5]

dim(wecare_phenotypes.df)
str(wecare_phenotypes.df)
wecare_phenotypes.df[1:5,1:5]

dim(wecare_variants.df)
colnames(wecare_variants.df)
wecare_variants.df[1:5,1:5]

dim(wecare_kgen.df)
colnames(wecare_kgen.df)
wecare_kgen.df[1:5,1:5]

dim(wecare_exac.df)
colnames(wecare_exac.df)
wecare_exac.df[1:5,1:5]

str(wecare.eigen)

sum(colnames(wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_variants.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_kgen.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_exac.df))

```

# function_to_calculate_eigenvectors

Implements procedure described by Price et al 2006 (PMID: 16862161)

```{r function_to_calculate_eigenvectors}

normalise_and_calculate_eigenvectors.udf <- function(x) {
  
  # --- Center and normalise variants (rows) --- #
  
  # Center by mean
  avg.rows <- apply(x, 1, mean, na.rm=TRUE)
  x.c <- x - avg.rows
  
  # Normalise by sqrt(p(1-p)) where p~"posterior estimate of unobserved allele frequency"
  # This is motivated by the fact that genetic drift per generation is proportional to this normalisation value (Patterson 2006)
  # Also this makes each column to have same variance
  # 
  p.fnc <- function(x) (1 + sum(x, na.rm=TRUE)) / (2 + 2 * sum(!is.na(x)))
  p <- apply(x, 1, p.fnc)
  eaf <- sqrt(p*(1-p))
  x.cn <- x.c/eaf
  
  # Substitute NAs to zeros
  0 -> x.cn[is.na(x)]
  
  # --- Calculate eigenvectors of covariance matrix of cases --- #
  
  cov.mx <- cov(x.cn)
  eig <- eigen(cov.mx) # eigenvectors in columns
  
  return(eig)

}

```

# calculate_total_AFs

Is used only for the common vars analysis below.  
Later AFs are recalculated after exclusion of eigenvectors outliers

```{r calculate_total_AFs}

# Calculate total wecare ac, an and af

ac_wecare_cln <- apply(wecare_genotypes.mx, 1, sum, na.rm=TRUE)
get_allele_number.udf <- function(x){2*sum(!is.na(x))}
an_wecare_cln <- apply(wecare_genotypes.mx, 1, get_allele_number.udf)
af_wecare_cln <- ac_wecare_cln/an_wecare_cln

# Ceck AFs 
# (note that uniform variants were excluded)
ac_wecare_cln[1:10]
an_wecare_cln[1:10]
af_wecare_cln[1:10]

min(ac_wecare_cln)
min(an_wecare_cln)
min(af_wecare_cln)

max(ac_wecare_cln)
max(an_wecare_cln)
max(af_wecare_cln)

rm(get_allele_number.udf, ac_wecare_cln, an_wecare_cln)

```

# make_subset_of_data_with_common_variants_only

```{r make_subset_of_data_with_common_variants_only}

# Note exclusion on both sides: high- and low- AFs
# Low AFs remove rare variants with common allele in reference genome
# Hight AFs remove rare variants with common allele in reference genome

wecare_common_vars <- af_wecare_cln > 0.05 & af_wecare_cln < 0.95
sum(wecare_common_vars) # 44,517
min(af_wecare_cln[wecare_common_vars])
max(af_wecare_cln[wecare_common_vars])

common_wecare_genotypes.mx <- wecare_genotypes.mx[wecare_common_vars,]
dim(common_wecare_genotypes.mx)
common_wecare_genotypes.mx[1:5,1:5]

common_wecare_variants.df <- wecare_variants.df[wecare_common_vars,]
dim(common_wecare_variants.df)
common_wecare_variants.df[1:5,1:5]

common_wecare_kgen.df <- wecare_kgen.df[wecare_common_vars,]
dim(common_wecare_kgen.df)
common_wecare_kgen.df[1:5,1:5]

common_wecare_exac.df <- wecare_exac.df[wecare_common_vars,]
dim(common_wecare_exac.df)
common_wecare_exac.df[1:5,1:5]

```

# calculate_and_plot_eigenvectors_wecare_common_variants

```{r calculate_and_plot_eigenvectors_wecare_common_variants}

# --- Calculate eigenvectors --- #

common_wecare.eigen <- normalise_and_calculate_eigenvectors.udf(common_wecare_genotypes.mx)

wecare_common_variants_eigenvectors.df <- as.data.frame(common_wecare.eigen$vectors) # eigenvectors in columns
wecare_common_variants_eigenvalues <- common_wecare.eigen$values

# --- Prepare data for plotting --- #

# Prepare colour scale
colours <- c("UBC" = "BLUE", "CBC" = "RED")
userColourScale <- scale_colour_manual(values=colours)

# Prepare cases lables
cases_labels <- as.vector(wecare_phenotypes.df$cc)
"CBC" -> cases_labels[cases_labels==1]
"UBC" -> cases_labels[cases_labels==0]

# Prepare cases IDs
cases_IDs <- as.vector(wecare_phenotypes.df$wes_id)

data2plot.df <- cbind(cases_IDs, cases_labels, wecare_common_variants_eigenvectors.df[,1:3])
colnames(data2plot.df) <- c("wes_id", "group", "ev1", "ev2", "ev3")

# --- Plot eig1 vs eig2 --- #

g <- ggplot(data2plot.df, aes(-ev1, ev2)) +
  geom_point(aes(colour=group, fill=group, text = wes_id)) + 
  labs(title="wecare common variants<br>(44,517 x 480)", x ="-eigenvector1", y = "eigenvector2") +
  userColourScale

ggplotly(g)

# --- Plot eig2 vs eig3 --- #

g <- ggplot(data2plot.df, aes(ev2, ev3)) +
  geom_point(aes(colour=group, fill=group, text = wes_id)) + 
  labs(title="wecare common variants<br>(44,517 x 480)", x ="eigenvector2", y = "eigenvector3") +
  userColourScale

ggplotly(g)

# --- Clean-up --- #

rm(wecare_common_variants_eigenvectors.df, wecare_common_variants_eigenvalues, g, data2plot.df, 
   cases_labels, cases_IDs, colours, userColourScale, wecare_common_vars, 
   af_wecare_cln, normalise_and_calculate_eigenvectors.udf)

```

# export_wecare_common_variants_to_EIGENSTRAT_format

```{r export_wecare_common_variants_to_EIGENSTRAT_format}

# --- make snp file --- #

# Extract data from variants table
wecare_common_variants_snp.df <- common_wecare_variants.df %>% 
  mutate(morgan = POS / 1000000) %>%  
  select(SplitVarID, CHROM, morgan, POS, REF, ALT)

# Recode CRHOM data
summary(wecare_common_variants_snp.df$CHROM)

str(wecare_common_variants_snp.df)
wecare_common_variants_snp.df$CHROM <- as.vector(wecare_common_variants_snp.df$CHROM)
str(wecare_common_variants_snp.df)

"23" -> wecare_common_variants_snp.df[wecare_common_variants_snp.df$CHROM == "X", "CHROM"]
"24" -> wecare_common_variants_snp.df[wecare_common_variants_snp.df$CHROM == "Y", "CHROM"]
"90" -> wecare_common_variants_snp.df[wecare_common_variants_snp.df$CHROM == "MT", "CHROM"]

wecare_common_variants_snp.df$CHROM <- as.factor(wecare_common_variants_snp.df$CHROM)
summary(wecare_common_variants_snp.df$CHROM)

# Write file (tab-separated)
write.table(wecare_common_variants_snp.df, 
            paste(interim_data_folder, "wecare_common_variants.snp", sep="/"), 
            quote=FALSE, sep="\t", 
            row.names=FALSE, col.names=FALSE) 

# --- make geno file --- #

# Get data from genotypes matrix
wecare_common_variants_geno.mx <- common_wecare_genotypes.mx

# Recode NA
9 -> wecare_common_variants_geno.mx[is.na(wecare_common_variants_geno.mx)] # NA is coded as 9

# Write file (no delimiters)
write.table(wecare_common_variants_geno.mx, 
            paste(interim_data_folder, "wecare_common_variants.geno", sep="/"),             
            quote=FALSE, sep="", 
            row.names=FALSE, col.names=FALSE)

# --- make ind file --- #
# has been made for wecare all variants

# Clean-up
rm(wecare_common_variants_snp.df, wecare_common_variants_geno.mx)

```

# compare_eigenvalues_and_eigenvectors_in_all_and_in_common_variants

```{r compare_eigenvalues_and_eigenvectors_in_all_and_in_common_variants}

# --- Compare eigenvalues --- #

eval_all <- wecare.eigen$values
eval_common <- common_wecare.eigen$values

plot(eval_all, main="Wecare eigenvalues (all variants)")
plot(eval_common, main="Wecare igenvalues (common variants)")
plot(eval_all,eval_common, main="Wecare eigenvalues (all vs common variants)")

# --- Compare eigenvectors --- #

# Gather data
ev1_all <- wecare.eigen$vectors[,1]
ev1_common <- common_wecare.eigen$vectors[,1]
ev2_all <- wecare.eigen$vectors[,2]
ev2_common <- common_wecare.eigen$vectors[,2]
ev3_all <- wecare.eigen$vectors[,3]
ev3_common <- common_wecare.eigen$vectors[,3]

data2plot.df <- as.data.frame(cbind(ev1_all, ev2_all, ev3_all, ev1_common, ev2_common, ev3_common))

# Calculate correlations
cor.test(ev1_all, ev1_common) # -0.482982, p-value = 2.2e-16
cor.test(ev2_all, ev2_common) # -0.2196824, p-value = 1.174e-06
cor.test(ev3_all, ev3_common) # -0.06442268, p-value = 0.1588

# Common sence check (these eigenvectors should be orthogonal...)
cor.test(ev1_all, ev2_all) # -3.530534e-16, p-value = 1
cor.test(ev1_common, ev2_common) # -6.391521e-16, p-value = 1

# Make plots
g <- ggplot(data2plot.df, aes(ev1_all, ev1_common)) +
  geom_point() + 
  labs(title="Wecare eigenvector 1<br>all vs common variants (p=2e-16)")
ggplotly(g)

g <- ggplot(data2plot.df, aes(ev2_all, ev2_common)) +
  geom_point() + 
  labs(title="Wecare eigenvector 2<br>all vs common variants (p=1e-06)")
ggplotly(g)

g <- ggplot(data2plot.df, aes(ev3_all, ev3_common)) +
  geom_point() + 
  labs(title="Wecare eigenvector 3<br>all vs common variants (p=0.16)")
ggplotly(g)

# Clean-up
rm(eval_all, eval_common, ev1_all, ev2_all, ev3_all, ev1_common, ev2_common, ev3_common, g, data2plot.df)

```

# data_summary

```{r data_summary}

ls()

dim(wecare_genotypes.mx)
class(wecare_genotypes.mx)
wecare_genotypes.mx[1:5,1:5]

dim(wecare_phenotypes.df)
str(wecare_phenotypes.df)
wecare_phenotypes.df[1:5,1:5]

dim(wecare_variants.df)
colnames(wecare_variants.df)
wecare_variants.df[1:5,1:5]

dim(wecare_kgen.df)
colnames(wecare_kgen.df)
wecare_kgen.df[1:5,1:5]

dim(wecare_exac.df)
colnames(wecare_exac.df)
wecare_exac.df[1:5,1:5]

str(wecare.eigen)

sum(colnames(wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_variants.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_kgen.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_exac.df))

# --- Subset with common variants only (AF>5%)

dim(common_wecare_genotypes.mx)
class(common_wecare_genotypes.mx)
common_wecare_genotypes.mx[1:5,1:5]

dim(common_wecare_variants.df)
colnames(common_wecare_variants.df)
common_wecare_variants.df[1:5,1:5]

dim(common_wecare_kgen.df)
colnames(common_wecare_kgen.df)
common_wecare_kgen.df[1:5,1:5]

dim(common_wecare_exac.df)
colnames(common_wecare_exac.df)
common_wecare_exac.df[1:5,1:5]

str(common_wecare.eigen)

sum(colnames(common_wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(common_wecare_genotypes.mx) != rownames(common_wecare_variants.df))
sum(rownames(common_wecare_genotypes.mx) != rownames(common_wecare_kgen.df))
sum(rownames(common_wecare_genotypes.mx) != rownames(common_wecare_exac.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r05a_calculate_egenvectors_wecare_only_jan2017_explore.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
