---
title: "calculate_eigenvectors_wecare_nfe_kgen50_jan2017"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 09Feb2017

# Summary

Overall, eigenvectors are calculated for 3 datasets:  

- wecare-only (all variants and common variants)
- wecare-nfe (all variants and common variants)  
- wecare-nfe-50kgen (overlapped-variants and overlapped-common-variants)  

This script deals with wecare-nfe-50kgen dataset 

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"

# Required libraries
library(dplyr)
library(ggplot2)
library(plotly)

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r01_read_and_clean_kgen50_jan2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(gt.mx)
class(gt.mx)
gt.mx[1:5, 1:5]

dim(vcf.df)
str(vcf.df)
vcf.df[1:5, 1:5]

dim(kgen50_cases.df)
str(kgen50_cases.df)
kgen50_cases.df[1:5,]

# Check consistence of rownames in gt.mx and vcf.df
sum(rownames(gt.mx) != rownames(vcf.df))

```

# function_to_calculate_eigenvectors

Implements procedure described by Price et al 2006 (PMID: 16862161)

```{r function_to_calculate_eigenvectors}

normalise_and_calculate_eigenvectors.udf <- function(x) {
  
  # --- Center and normalise variants (rows) --- #
  
  # Center by mean
  avg.rows <- apply(x, 1, mean, na.rm=TRUE)
  x.c <- x - avg.rows
  
  # Normalise by sqrt(p(1-p)) where p~"posterior estimate of unobserved allele frequency"
  # This is motivated by the fact that genetic drift per generation is proportional to this normalisation value (Patterson 2006)
  # Also this makes each column to have same variance
  # 
  p.fnc <- function(x) (1 + sum(x, na.rm=TRUE)) / (2 + 2 * sum(!is.na(x)))
  p <- apply(x, 1, p.fnc)
  eaf <- sqrt(p*(1-p))
  x.cn <- x.c/eaf
  
  # Substitute NAs to zeros
  0 -> x.cn[is.na(x)]
  
  # --- Calculate eigenvectors of covariance matrix of cases --- #
  
  cov.mx <- cov(x.cn)
  eig <- eigen(cov.mx) # eigenvectors in columns
  
  return(eig)

}

```

# calculate_and_plot_eigenvectors_full_data

```{r calculate_and_plot_eigenvectors_full_data}

# --- Calculate eigenvectors --- #

wecare_nfe_kgen50_eigen <- normalise_and_calculate_eigenvectors.udf(gt.mx)

evectors.df <- as.data.frame(wecare_nfe_kgen50_eigen$vectors) # eigenvectors in columns
evalues <- wecare_nfe_kgen50_eigen$values

# --- Prepare data for plotting --- #

# Prepare cases IDs
cases_IDs <- colnames(gt.mx)

# Prepare cases lables
kgen50_populations <- as.vector(kgen50_cases.df[colnames(gt.mx)[1:50],"V3"])
cases_labels <- c(kgen50_populations, rep("NFE",198), rep("WECARE",480))

# make the dataframe
data2plot.df <- cbind(cases_IDs, cases_labels, evectors.df[,1:3])
colnames(data2plot.df) <- c("sample", "group", "ev1", "ev2", "ev3")

# Prepare colour scale
colours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", "SAS" = "BLACK", "EAS" = "MAGENTA", "NFE" = "PINK", "WECARE" = "RED")
userColourScale <- scale_colour_manual(values=colours)

# --- Plot eig1 vs eig2 --- #

g <- ggplot(data2plot.df, aes(-ev1, ev2)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>all overlapped variants (83,881 x 728)", x ="-eigenvector1", y = "eigenvector2") +
  userColourScale

ggplotly(g)

# --- Plot eig2 vs eig3 --- #

g <- ggplot(data2plot.df, aes(ev2, ev3)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>all overlapped variants (83,881 x 728)", x ="eigenvector2", y = "eigenvector3") +
  userColourScale

ggplotly(g)

# --- Clean-up --- #

rm(evectors.df, evalues, colours, g, data2plot.df, kgen50_populations)

```

# calculate_and_plot_eigenvectors_common_variants

```{r calculate_and_plot_eigenvectors_common_variants}

# --- Calculate AFs --- #

ac_cln <- apply(gt.mx, 1, sum, na.rm=TRUE)
get_allele_number.fnc <- function(x){2*sum(!is.na(x))}
an_cln <- apply(gt.mx, 1, get_allele_number.fnc)
af_cln <- ac_cln/an_cln

# Ceck AFs 
# (note that uniform variants were excluded)
ac_cln[1:10]
an_cln[1:10]
af_cln[1:10]

min(ac_cln)
min(an_cln)
min(af_cln)

max(ac_cln)
max(an_cln)
max(af_cln)

# Add updated AFs to vcf.df
vcf.df <- cbind(vcf.df, ac_cln, an_cln, af_cln)

# --- Exclude rare variants --- #
# Note exclusion on both sides: high- and low- AFs
# Low AFs remove rare variants with common allele in reference genome
# Hight AFs remove rare variants with common allele in reference genome

common_vars <- af_cln > 0.05 & af_cln < 0.95
sum(common_vars) # 43,563
min(af_cln[common_vars])
max(af_cln[common_vars])

common_gt.mx <- gt.mx[common_vars,]
common_vcf.df <- vcf.df[common_vars,]
dim(common_gt.mx)
dim(common_vcf.df)

# --- Calculate eigenvectors --- #

common_wecare_nfe_kgen50_eigen <- normalise_and_calculate_eigenvectors.udf(common_gt.mx)

common_evectors.df <- as.data.frame(common_wecare_nfe_kgen50_eigen$vectors) # eigenvectors in columns
common_evalues <- common_wecare_nfe_kgen50_eigen$values

# --- Prepare data for plotting --- #

# make the dataframe
common_data2plot.df <- cbind(cases_IDs, cases_labels, common_evectors.df[,1:3])
colnames(common_data2plot.df) <- c("sample", "group", "ev1", "ev2", "ev3")

# --- Plot eig1 vs eig2 --- #

g <- ggplot(common_data2plot.df, aes(-ev1, -ev2)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>overlapped common vars (43,563 x 728)", x ="-eigenvector1", y = "-eigenvector2") +
  userColourScale

ggplotly(g)

# --- Plot eig2 vs eig3 --- #

g <- ggplot(common_data2plot.df, aes(ev2, ev3)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>overlapped common vars (43,563 x 728)", x ="eigenvector2", y = "eigenvector3") +
  userColourScale

ggplotly(g)

# --- Clean-up --- #

rm(ac_cln, an_cln, af_cln, get_allele_number.fnc, common_vars, common_evectors.df, common_evalues, cases_IDs, cases_labels, userColourScale, g, normalise_and_calculate_eigenvectors.udf, common_data2plot.df)

```

# compare_eigenvalues_and_eigenvectors_in_all_and_common_variants

```{r compare_eigenvalues_and_eigenvectors_in_all_and_common_variants}

# --- Compare eigenvalues --- #

eval_all <- wecare_nfe_kgen50_eigen$values
eval_common <- common_wecare_nfe_kgen50_eigen$values

plot(eval_all, main="Wecare-nfe-kgen50 eigenvalues (all variants)")
plot(eval_common, main="Wecare-nfe-kgen50 eigenvalues (common variants)")
plot(eval_all,eval_common, main="Wecare-nfe-kgen50 eigenvalues (all vs common variants)")

# --- Compare eigenvectors --- #

# Gather data
ev1_all <- wecare_nfe_kgen50_eigen$vectors[,1]
ev1_common <- common_wecare_nfe_kgen50_eigen$vectors[,1]
ev2_all <- wecare_nfe_kgen50_eigen$vectors[,2]
ev2_common <- common_wecare_nfe_kgen50_eigen$vectors[,2]
ev3_all <- wecare_nfe_kgen50_eigen$vectors[,3]
ev3_common <- common_wecare_nfe_kgen50_eigen$vectors[,3]

data2plot.df <- as.data.frame(cbind(ev1_all, ev2_all, ev3_all, ev1_common, ev2_common, ev3_common))

# Calculate correlations
cor.test(ev1_all, ev1_common) # 0.8278741, p-value < 2.2e-16
cor.test(ev2_all, ev2_common) # -0.5837235, p-value < 2.2e-16
cor.test(ev3_all, ev3_common) # -0.003093386, p-value = 0.9336

# Common sence check (these eigenvectors should be orthogonal...)
cor.test(ev1_all, ev2_all) # 2.15652e-16, p-value = 1
cor.test(ev1_common, ev2_common) # 2.692919e-16, p-value = 1

# Make plots
g <- ggplot(data2plot.df, aes(ev1_all, ev1_common)) +
  geom_point() + 
  labs(title="Wecare-nfe-kgen50 eigenvector 1<br>all vs common variants (p=0.0003)")
ggplotly(g)

g <- ggplot(data2plot.df, aes(ev2_all, ev2_common)) +
  geom_point() + 
  labs(title="Wecare-nfe-kgen50 eigenvector 2<br>all vs common variants (p=8e-13)")
ggplotly(g)

g <- ggplot(data2plot.df, aes(ev3_all, ev3_common)) +
  geom_point() + 
  labs(title="Wecare-nfe-kgen50 eigenvector 3<br>all vs common variants (p-value = 0.2193)")
ggplotly(g)

# Clean-up
rm(eval_all, eval_common, ev1_all, ev2_all, ev3_all, ev1_common, ev2_common, ev3_common, g, data2plot.df)

```

# export_wecare_nfe_khen50_to_EIGENSTRAT_format
Omitted

# data_summary

```{r data_summary}

ls()

# wecare nfe

dim(gt.mx)
class(gt.mx)
gt.mx[1:5,1:5]

dim(common_gt.mx)
class(common_gt.mx)
common_gt.mx[1:5,1:5]

dim(vcf.df)
str(vcf.df)
vcf.df[1:5,1:5]

dim(common_vcf.df)
str(common_vcf.df)
common_vcf.df[1:5,1:5]

dim(kgen50_cases.df)
colnames(kgen50_cases.df)
kgen50_cases.df[1:5,]

class(wecare_nfe_kgen50_eigen)
str(wecare_nfe_kgen50_eigen)

class(common_wecare_nfe_kgen50_eigen)
str(common_wecare_nfe_kgen50_eigen)

sum(rownames(gt.mx) != rownames(vcf.df))
sum(rownames(common_gt.mx) != rownames(common_vcf.df))
sum(colnames(gt.mx) != colnames(common_gt.mx))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r05c_calculate_egenvectors_wecare_nfe_kgen50_jan2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
