---
title: "exclude_eigenvectors_outliers_wecare_nfe_jan2017"
output: html_document
---

started: Alexey Larionov, 16Feb2017  
last updated: Alexey Larionov, 01Mar2017

# Summary

Exclude eigenvectors outliers and clean data for wecare nfe jan2017 dataset

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"
results_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/results"

```

# read_data

```{r read_data}

load(paste(interim_data_folder, "r05b_calculate_egenvectors_wecare_nfe_jan2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5, 1:5]

dim(wecare_nfe_variants.df)
str(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5, 1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
str(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
str(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

str(wecare_nfe.eigen)

# Check consistence of rownames in gt.mx and vcf.df
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))
sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))

```

# add_wes_eigenvectors_to_phenotypes_table

```{r add_wes_eigenvectors_to_phenotypes_table}

wecare_nfe_phenotypes_cols <- colnames(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df <- cbind(wecare_nfe_phenotypes.df, wecare_nfe.eigen$vectors[,1:3])
colnames(wecare_nfe_phenotypes.df) <- c(wecare_nfe_phenotypes_cols, "eig1_wes", "eig2_wes", "eig3_wes")

rm(wecare_nfe_phenotypes_cols, wecare_nfe.eigen)

```

# remove_wes_eigenvectors_outliers 

The outliers have been selected earlier:  
P6_D05, P5_E09, P1_C12, P1_E06, P1_D08, P1_C05

```{r remove_wes_eigenvectors_outliers}

# Flag outliers
eig_outliers=c("P6_D05", "P5_E09", "P1_C12", "P1_E06", "P1_D08", "P1_C05")
"eigenvector_outlier" -> wecare_nfe_phenotypes.df[wecare_nfe_phenotypes.df$wes_id %in% eig_outliers,"filter"]

# Count outliers
outliers <- wecare_nfe_phenotypes.df$filter == "eigenvector_outlier"
sum(outliers)

# Remove outliers
wecare_nfe_phenotypes.df <- wecare_nfe_phenotypes.df[!outliers,]
wecare_nfe_genotypes.mx <- wecare_nfe_genotypes.mx[,!outliers]

# Clean-up
rm(eig_outliers, outliers)

```

# remove_variants_with_uniform_genotypes_accross_all_samples
Remove 4,078 variants: 275,516 -> 271,438

```{r remove_variants_with_uniform_genotypes_accross_all_samples}

# Check that there is no all-NA variants
non_NA_count.udf <- function(x){sum(!is.na(x))}
all_NA <- apply(wecare_nfe_genotypes.mx, 1, non_NA_count.udf) == 0
sum(all_NA) # 0

# Function to detect uniform numeric vector
uniform_vector.udf <- function(x){
  if(min(x, na.rm=TRUE) == max(x, na.rm=TRUE)){return(TRUE)} else {return(FALSE)}}

# Variants with uniform genotypes accross all samples 
uniform_genotypes <- apply(wecare_nfe_genotypes.mx, 1, uniform_vector.udf)
summary(uniform_genotypes)
sum(uniform_genotypes) # 4,078

# Remove variants with uniform genotypes accross all samples
wecare_nfe_genotypes.mx <- wecare_nfe_genotypes.mx[!uniform_genotypes,]
dim(wecare_nfe_genotypes.mx) # 271438    672
wecare_nfe_genotypes.mx[1:5,1:5]

wecare_nfe_variants.df <- wecare_nfe_variants.df[!uniform_genotypes,]
dim(wecare_nfe_variants.df) # 271438     26
wecare_nfe_variants.df[1:5,1:5]

wecare_nfe_kgen.df <- wecare_nfe_kgen.df[!uniform_genotypes,]
dim(wecare_nfe_kgen.df) # 271438      9
wecare_nfe_kgen.df[1:5,1:5]

wecare_nfe_exac.df <- wecare_nfe_exac.df[!uniform_genotypes,]
dim(wecare_nfe_exac.df) # 271438     48
wecare_nfe_exac.df[1:5,1:5]

# Clean-up
rm(non_NA_count.udf, all_NA, uniform_vector.udf, uniform_genotypes)

```

# explore_remaining_cases

232 CBC, 242 UBC, 198 NFE

```{r explore_remaining_cases}

# Numbers of cases and controls
summary(as.factor(wecare_nfe_phenotypes.df$cc))

```

# calculate_total_AFs

```{r calculate_total_AFs}

# Function to count alleles (2x for diploid)
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Check genotypes table
dim(wecare_nfe_genotypes.mx)

# Calculate ac, an and af
ac_wecare_nfe_cln <- apply(wecare_nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_wecare_nfe_cln <- apply(wecare_nfe_genotypes.mx, 1, get_allele_number.udf)
af_wecare_nfe_cln <- ac_wecare_nfe_cln/an_wecare_nfe_cln

# Ceck AFs 
# (note that uniform variants were excluded)
ac_wecare_nfe_cln[1:6]
an_wecare_nfe_cln[1:6]
af_wecare_nfe_cln[1:6]

min(ac_wecare_nfe_cln)
min(an_wecare_nfe_cln)
min(af_wecare_nfe_cln)

max(ac_wecare_nfe_cln)
max(an_wecare_nfe_cln)
max(af_wecare_nfe_cln)

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_wecare_nfe_cln, an_wecare_nfe_cln, af_wecare_nfe_cln)

# Clean-up
rm(ac_wecare_nfe_cln, an_wecare_nfe_cln, af_wecare_nfe_cln)

```

# calculate_nfe_AFs

```{r calculate_nfe_AFs}

# Prepare genotypes table
nfe_cases <- wecare_nfe_phenotypes.df$cc == -1 
sum(nfe_cases) # 198
nfe_genotypes.mx <- wecare_nfe_genotypes.mx[,nfe_cases]
dim(nfe_genotypes.mx)

# Calculate nfe ac, an and af
ac_nfe_cln <- apply(nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_nfe_cln <- apply(nfe_genotypes.mx, 1, get_allele_number.udf)
af_nfe_cln <- ac_nfe_cln/an_nfe_cln

# Ceck AFs 
ac_nfe_cln[1:6]
an_nfe_cln[1:6]
af_nfe_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_nfe_cln, an_nfe_cln, af_nfe_cln)

# Clean-up
rm(nfe_cases, nfe_genotypes.mx, ac_nfe_cln, an_nfe_cln, af_nfe_cln)

```

# calculate_wecare_AFs

```{r calculate_wecare_AFs}

# Prepare genotypes table
wecare_cases <- wecare_nfe_phenotypes.df$cc == 0 | wecare_nfe_phenotypes.df$cc == 1
sum(wecare_cases) # 474
wecare_genotypes.mx <- wecare_nfe_genotypes.mx[,wecare_cases]
dim(wecare_genotypes.mx)

# Calculate nfe ac, an and af
ac_wecare_cln <- apply(wecare_genotypes.mx, 1, sum, na.rm=TRUE)
an_wecare_cln <- apply(wecare_genotypes.mx, 1, get_allele_number.udf)
af_wecare_cln <- ac_wecare_cln/an_wecare_cln

# Ceck AFs 
ac_wecare_cln[1:6]
an_wecare_cln[1:6]
af_wecare_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_wecare_cln, an_wecare_cln, af_wecare_cln)

# Clean-up
rm(wecare_cases, wecare_genotypes.mx, ac_wecare_cln, an_wecare_cln, af_wecare_cln)

```

# calculate_ubc_AFs

```{r calculate_ubc_AFs}

# Prepare genotypes table
ubc_cases <- wecare_nfe_phenotypes.df$cc == 0 
sum(ubc_cases) # 242
ubc_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_cases]
dim(ubc_genotypes.mx)

# Calculate ubc ac, an and af
ac_ubc_cln <- apply(ubc_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_cln <- apply(ubc_genotypes.mx, 1, get_allele_number.udf)
af_ubc_cln <- ac_ubc_cln/an_ubc_cln

# Ceck AFs 
ac_ubc_cln[1:6]
an_ubc_cln[1:6]
af_ubc_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_ubc_cln, an_ubc_cln, af_ubc_cln)

# Clean-up
rm(ubc_cases, ubc_genotypes.mx, ac_ubc_cln, an_ubc_cln, af_ubc_cln)

```

# calculate_cbc_AFs

```{r calculate_cbc_AFs}

# Prepare genotypes table
cbc_cases <- wecare_nfe_phenotypes.df$cc == 1 
sum(cbc_cases) # 232
cbc_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_cases]
dim(cbc_genotypes.mx)

# Calculate cbc ac, an and af
ac_cbc_cln <- apply(cbc_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_cln <- apply(cbc_genotypes.mx, 1, get_allele_number.udf)
af_cbc_cln <- ac_cbc_cln/an_cbc_cln

# Ceck AFs 
ac_cbc_cln[1:6]
an_cbc_cln[1:6]
af_cbc_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_cbc_cln, an_cbc_cln, af_cbc_cln)

# Clean-up
rm(cbc_cases, cbc_genotypes.mx, ac_cbc_cln, an_cbc_cln, af_cbc_cln)

```

# calculate_ubc_fam_AFs

```{r calculate_ubc_fam_AFs}

# Prepare genotypes table
ubc_fam_cases <- wecare_nfe_phenotypes.df$cc == 0 & wecare_nfe_phenotypes.df$family_history == 1
sum(ubc_fam_cases) # 70
ubc_fam_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_fam_cases]
dim(ubc_fam_genotypes.mx)

# Calculate ubc_fam ac, an and af
ac_ubc_fam_cln <- apply(ubc_fam_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_fam_cln <- apply(ubc_fam_genotypes.mx, 1, get_allele_number.udf)
af_ubc_fam_cln <- ac_ubc_fam_cln/an_ubc_fam_cln

# Ceck AFs 
ac_ubc_fam_cln[1:6]
an_ubc_fam_cln[1:6]
af_ubc_fam_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_ubc_fam_cln, an_ubc_fam_cln, af_ubc_fam_cln)

# Clean-up
rm(ubc_fam_cases, ubc_fam_genotypes.mx, ac_ubc_fam_cln, an_ubc_fam_cln, af_ubc_fam_cln)

```

# calculate_ubc_nofam_AFs

```{r calculate_ubc_nofam_AFs}

# Prepare genotypes table
ubc_nofam_cases <- wecare_nfe_phenotypes.df$cc == 0 & wecare_nfe_phenotypes.df$family_history == 0
sum(ubc_nofam_cases) # 172
ubc_nofam_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_nofam_cases]
dim(ubc_nofam_genotypes.mx)

# Calculate ubc_nofam ac, an and af
ac_ubc_nofam_cln <- apply(ubc_nofam_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_nofam_cln <- apply(ubc_nofam_genotypes.mx, 1, get_allele_number.udf)
af_ubc_nofam_cln <- ac_ubc_nofam_cln/an_ubc_nofam_cln

# Ceck AFs 
ac_ubc_nofam_cln[1:6]
an_ubc_nofam_cln[1:6]
af_ubc_nofam_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_ubc_nofam_cln, an_ubc_nofam_cln, af_ubc_nofam_cln)

# Clean-up
rm(ubc_nofam_cases, ubc_nofam_genotypes.mx, ac_ubc_nofam_cln, an_ubc_nofam_cln, af_ubc_nofam_cln)

```

# calculate_cbc_fam_AFs

```{r calculate_cbc_fam_AFs}

# Prepare genotypes table
cbc_fam_cases <- wecare_nfe_phenotypes.df$cc == 1 & wecare_nfe_phenotypes.df$family_history == 1
sum(cbc_fam_cases) # 82
cbc_fam_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_fam_cases]
dim(cbc_fam_genotypes.mx)

# Calculate cbc_fam ac, an and af
ac_cbc_fam_cln <- apply(cbc_fam_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_fam_cln <- apply(cbc_fam_genotypes.mx, 1, get_allele_number.udf)
af_cbc_fam_cln <- ac_cbc_fam_cln/an_cbc_fam_cln

# Ceck AFs 
ac_cbc_fam_cln[1:6]
an_cbc_fam_cln[1:6]
af_cbc_fam_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_cbc_fam_cln, an_cbc_fam_cln, af_cbc_fam_cln)

# Clean-up
rm(cbc_fam_cases, cbc_fam_genotypes.mx, ac_cbc_fam_cln, an_cbc_fam_cln, af_cbc_fam_cln)

```

# calculate_cbc_nofam_AFs

```{r calculate_cbc_nofam_AFs}

# Prepare genotypes table
cbc_nofam_cases <- wecare_nfe_phenotypes.df$cc == 1 & wecare_nfe_phenotypes.df$family_history == 0
sum(cbc_nofam_cases) # 150
cbc_nofam_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_nofam_cases]
dim(cbc_nofam_genotypes.mx)

# Calculate cbc_nofam ac, an and af
ac_cbc_nofam_cln <- apply(cbc_nofam_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_nofam_cln <- apply(cbc_nofam_genotypes.mx, 1, get_allele_number.udf)
af_cbc_nofam_cln <- ac_cbc_nofam_cln/an_cbc_nofam_cln

# Ceck AFs 
ac_cbc_nofam_cln[1:6]
an_cbc_nofam_cln[1:6]
af_cbc_nofam_cln[1:6]

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, 
  ac_cbc_nofam_cln, an_cbc_nofam_cln, af_cbc_nofam_cln)

# Clean-up
rm(cbc_nofam_cases, cbc_nofam_genotypes.mx, ac_cbc_nofam_cln, 
   an_cbc_nofam_cln, af_cbc_nofam_cln, get_allele_number.udf)

```

# data_summary

```{r data_summary}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_variants.df)
colnames(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))

sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r06b_exclude_egenvectors_outliers_wecare_nfe_jan2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
