---
title: "filter_by_variant_effect_wecare_only_jan2017_explore"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 24Feb2017

# Summary

### Exploring removal of common (AF>5-10%) variants
Modest effect. Slightly playes down MUC4 etc. 

###Exploring BRCA1, BRCA2, PALB2 carriers
Still there is a number of BRCA1, BRCA2, PALB2 carriers.  
All detected variants were reviewed in Dec 2016 and found non-pathogenic. 

Interestingly, there was  
- 2 compound hets (BRCA1 and BRCA2, 39 and 43 yrs UBC) and  
- 3 cases of BRCA1+BRCA2 and 1 case of BRCA1+PALB2 (all CBC of 40-50 yrs)

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_01.17/scripts")
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_01.17/interim_data"

# Load libraries
library(dplyr)

```

# load_data

```{r load_data}

# Load data
load(paste(interim_data_folder, "s07a_filter_by_variant_effect_wecare_only_jan2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_genotypes.mx)
class(wecare_genotypes.mx)
wecare_genotypes.mx[1:5,1:5]

dim(wecare_phenotypes.df)
str(wecare_phenotypes.df)
wecare_phenotypes.df[1:5,1:5]

dim(wecare_variants.df)
colnames(wecare_variants.df)
wecare_variants.df[1:5,1:5]

dim(wecare_kgen.df)
colnames(wecare_kgen.df)
wecare_kgen.df[1:5,1:5]

dim(wecare_exac.df)
colnames(wecare_exac.df)
wecare_exac.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_variants.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_kgen.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_exac.df))

```

# explore_non_common_variants

Defining variants that are not common as the following:  
[ kgen EUR AF < 5% or NA ] AND [ wecare AF < 10 ]

This overall changes are modest:
the number of variants drops from 17,614 to 16,515
the number of genes drops from 8,770 to 8,478
It seems that the upstream selection implicitly removed most of common variants. 

Weighting by AF in the downstream analysis should take care of the remaining ~5% of proteing-affecting variants that happened to be common. 

Genes with diverse non-common protein-affecting variants still include:
ATM (14 types of variants), POLG (12), POLQ(11), BRCA2 (10)
Other potentially interesting genes with many different types of protein-affecting variant include:
FN1 (10), MKI67 (8), CHEK2 (7), MLH3 (7), 

```{r explore_non_common_variants}

# Non-common  kgen
non_common_kgen_variants <- wecare_kgen.df$kgen.EUR_AF < 0.05
summary(non_common_kgen_variants)
TRUE -> non_common_kgen_variants[is.na(non_common_kgen_variants)]
summary(non_common_kgen_variants) # 16,601

# Non-common wecare
non_common_wecare_variants <- wecare_variants.df$af_wecare_cln < 0.1
summary(non_common_wecare_variants) # 16,824

# Non-common combined
non_common_variants <- non_common_kgen_variants & non_common_wecare_variants
summary(non_common_variants) # 16,515

# Get number of retained genes
length(unique(wecare_variants.df[non_common_variants,"SYMBOL"])) # 8,478

# Variants per gene
x <- sort(table(as.vector(wecare_variants.df[non_common_variants,"SYMBOL"])), decreasing = TRUE)
x[1:300]

# Save table of counts
y <- cbind(names(x),x)
c("gene","count") -> colnames(y)
write.table(y, file=paste(results_folder, "wecare_only_non_common_variants_per_gene.txt", sep="/"), quote = FALSE, row.names = FALSE)

# Plot histogram
z <- wecare_variants.df %>%  group_by(SYMBOL) %>% summarise(n())
"vars_count" -> colnames(z)[2]
hist(z$vars_count, main="Histogram for numbers of non-common protein-affecting variants per gene", xlab="Variants per gene", ylab="Number of genes", ylim=c(0,7000), breaks=25, labels = TRUE)
mean(z$vars_count) # 2.008438

# Clean-up
rm(non_common_kgen_variants, non_common_wecare_variants, non_common_variants, x,y,z)

```

# explore_BRCA1_carriers

4 different BRCA1 protein-affecting variants:  
all missenses (one pretty common: ~5%),  
all reviuved and considered non-pathogenic in Dec2016.  

52 BRCA1 carriers (mainly carrying the common variant). 

One compund heterozygous case with two rare (<0.1% each) "non-pathogenic" missences:  
P1_H07 carrying rs56012641 & rs56187033: early onset (39yrs) unilateral BC

```{r explore_BRCA1_carriers}

# Select variants
brca1_variants <- wecare_variants.df$SYMBOL == "BRCA1"
summary(brca1_variants) # 4

# Select carriers
brca1_count_per_case <- apply(wecare_genotypes.mx[brca1_variants,], 2, sum, na.rm=TRUE)
summary(as.factor(brca1_count_per_case))
brca1_carrier <- brca1_count_per_case > 0
summary(brca1_carrier) # 52

# Subset dataset
wecare_brca1_genotypes.mx <- wecare_genotypes.mx[brca1_variants, brca1_carrier]
wecare_brca1_exac.df <- wecare_exac.df[brca1_variants,]
wecare_brca1_kgen.df <- wecare_kgen.df[brca1_variants,]
wecare_brca1_variants.df <- wecare_variants.df[brca1_variants,]
wecare_brca1_phenotypes.df <- wecare_phenotypes.df[brca1_carrier,]

# rs-ids: all have been revieved in Dec2016 and considered non-pathogenic. 
wecare_brca1_variants.df[,c("SplitVarID","Existing_variation")]

# One case with compond heterozygous: P1_H07 carrying rs56012641 & rs56187033
brca1_homs <- apply(wecare_brca1_genotypes.mx, 2, sum, na.rm=TRUE) == 2
wecare_brca1_genotypes.mx[,brca1_homs]


wecare_phenotypes.df["P1_H07",c("cc","age_dx","age_ref","rstime")]

```

# explore_BRCA2_carriers

10 diferent BRCA2 protein-affecting variants:  
9 dd-misseswses and 1 relatively common stop-gain (~1%).  
All reviuved in Dec2016 and considered non-pathogenic.  

30 BRCA2 carriers: most carrying the common stop-gain.   

One compund heterozygous case carrying the common stop-gain and a rare dd missence:  
P3_C12 carrying rs11571747 & rs11571833 (UBC at 43 yrs)

```{r explore_BRCA2_carriers}

# Select variants
brca2_variants <- wecare_variants.df$SYMBOL == "BRCA2"
sum(brca2_variants) # 10

# Select carriers
brca2_count_per_case <- apply(wecare_genotypes.mx[brca2_variants,], 2, sum, na.rm=TRUE)
summary(as.factor(brca2_count_per_case))
brca2_carrier <- brca2_count_per_case > 0
summary(brca2_carrier) # 30

# Subset dataset
wecare_brca2_genotypes.mx <- wecare_genotypes.mx[brca2_variants, brca2_carrier]
wecare_brca2_exac.df <- wecare_exac.df[brca2_variants,]
wecare_brca2_kgen.df <- wecare_kgen.df[brca2_variants,]
wecare_brca2_variants.df <- wecare_variants.df[brca2_variants,]
wecare_brca2_phenotypes.df <- wecare_phenotypes.df[brca2_carrier,]

# rs-ids: all have been revieved in Dec2016 and considered non-pathogenic. 
wecare_brca2_variants.df[,c("SplitVarID","Existing_variation")]

# One case with compond heterozygous: P3_C12 carrying rs11571747 & rs11571833
# the first of the variants is a rare dd missence, the second is a relatively common stop-gain (~1%)
brca2_homs <- apply(wecare_brca2_genotypes.mx, 2, sum, na.rm=TRUE) == 2
wecare_brca2_genotypes.mx[,brca2_homs]
colnames(wecare_brca2_genotypes.mx)[brca2_homs]

wecare_phenotypes.df["P3_C12",c("cc","age_dx","age_ref","rstime")]

```

# explore_PALB2_carriers

5 diferent PALB2 protein-affecting variants: all dd-missenses, one - pretty common (~2.5% AF).  
All reviuved in Dec2016 and considered non-pathogenic.  

34 PALB2 carriers, most carrying the common missense.  

```{r explore_PALB2_carriers}

# Select variants
palb2_variants <- wecare_variants.df$SYMBOL == "PALB2"
sum(palb2_variants) # 5

# Select carriers
palb2_count_per_case <- apply(wecare_genotypes.mx[palb2_variants,], 2, sum, na.rm=TRUE)
summary(as.factor(palb2_count_per_case)) # No homs or compond hets
palb2_carrier <- palb2_count_per_case > 0
summary(palb2_carrier) # 34

# Subset dataset
wecare_palb2_genotypes.mx <- wecare_genotypes.mx[palb2_variants, palb2_carrier]
wecare_palb2_exac.df <- wecare_exac.df[palb2_variants,]
wecare_palb2_kgen.df <- wecare_kgen.df[palb2_variants,]
wecare_palb2_variants.df <- wecare_variants.df[palb2_variants,]
wecare_palb2_phenotypes.df <- wecare_phenotypes.df[palb2_carrier,]

# rs-ids: all have been revieved in Dec2016 and considered non-pathogenic. 
wecare_palb2_variants.df[,c("SplitVarID","Existing_variation")]

# Cleanup
#rm(palb2_variants, palb2_count_per_case, palb2_carrier, 
#   wecare_palb2_genotypes.mx, wecare_palb2_exac.df, wecare_palb2_kgen.df, 
#   wecare_palb2_variants.df, wecare_palb2_phenotypes.df)

```

# brca1_brca2_combinations

3 cases simulteneously carrying BRCA1 and BRCA2 variants:

2 (P2_D05, P5_D08): carry het common (6%) dd-missense BRCA1 + het common (1%) stop-gain BRCA2
1 (P5_D05): carry hom common (0.36%=6%*6%) dd-missense BRCA1 + het rare (0.5%) dd-missense BRCA2

It seems that such number of combinations of non-pathogenic BRCA1-BRCA2 variants (at least for the 1st combination) would be possible to expect by chance.  However, in each of 3 cases this "chance" coinsided with early onset (40-50 yrs) bilateral bc. 

```{r brca1_brca2_combinations}

brca1_brca2_carrier <- brca1_carrier & brca2_carrier
sum(brca1_brca2_carrier) # 3

brca1_brca2_carrier <- wecare_phenotypes.df[brca1_brca2_carrier,"wes_id"]
wecare_brca1_genotypes.mx[,brca1_brca2_carrier]
wecare_brca2_genotypes.mx[,brca1_brca2_carrier]

wecare_phenotypes.df[c("P2_D05", "P5_D08", "P5_D05"),c("cc","age_dx","age_ref","rstime")]

```

# brca1_palb2_combinations

1 case (P4_C12) simulteneously carryes BRCA1 and PALB2 variants:
het common (6%) dd-missense BRCA1 + het common (3%) dd-missense PALB2.

One such combination is likely to happen in 500 cases by chance (6%*3%/512).
However, this chance coinsided with early onset (41 yrs) bilateral (41 + 45 yrs) bc phenotype. 

```{r brca1_palb2_combinations}

brca1_palb2_carrier <- brca1_carrier & palb2_carrier
sum(brca1_palb2_carrier) # 1

brca1_palb2_carrier <- wecare_phenotypes.df[brca1_palb2_carrier,"wes_id"]
brca1_palb2_carrier
wecare_brca1_genotypes.mx[,brca1_palb2_carrier]
wecare_palb2_genotypes.mx[,brca1_palb2_carrier]

wecare_phenotypes.df[brca1_palb2_carrier,c("cc","age_dx","age_ref","rstime")]

```

# brca2_palb2_combinations
none

```{r brca2_palb2_combinations}

brca2_palb2_carrier <- brca2_carrier & palb2_carrier
sum(brca2_palb2_carrier) # 0

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "s07a_filter_by_variant_effect_wecare_only_jan2017_explore.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
